Understanding models for spatial structure in small-area estimation

### 2023-05-31 src/docs_paper/paper.Rmd
Paper writing and update template

---
title: Understanding models for spatial structure in small-area estimation
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A,B
    corresponding: ath19@ic.ac.uk
  - name: Jeffrey W. Eaton
    affiliation: C,B
  - name: Seth R. Flaxman
    affiliation: D
address:
  - code: A
    address: Department of Mathematics, Imperial College London
  - code: B
    address: MRC Centre for Global Infectious Disease Analysis, Imperial College London
  - code: C
    address: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health, Harvard University
  - code: D
    address: Department of Computer Science, University of Oxford
abstract: |
  |  Spatial correlation between small-areas is typically accounted for within a generalised linear mixed model setting using spatial random effects. However, commonly used models for spatial random effects based upon adjacency relations, like the Besag model, make unrealistic and unsatisfying spatial assumptions when the geometry is not a grid. Furthermore, they do not take into account that across many settings areal data is generated by aggregating continuous point data. Defining spatial random effects which instead correspond to aggregated Gaussian processes gives a more intuitively convincing correlation structure, and allows tools from kernel methods to be used in models for areal data.
  |
  |  In a simulation study, we used a proper scoring rule to evaluate the performance of different spatial random effect specifications. The simulations were performed under model misspecification and across a range of vignette and realistic geometries with different regularities. We also compared model performance in estimating district-level HIV prevalence from PHIA household survey data across four countries in sub-Saharan Africa, using cross-validation and spatial cross-validation.

author_summary: |
  Author summary

bibliography: citations.bib
classoption: preprint, authoryear
output:
  rticles::elsevier_article:
    number_sections: true
    includes:
      in_header: preamble.tex
csl: elsarticle.cls
---

@@

Spatial random effects are frequently used to model spatial variation in areal data [@haining2003spatial; @cramb2018investigation].
Gaussian Markov random fields (GMRFs) [@rue2005gaussian], which combine a Gaussian distribution with Markov conditional independence assumptions between areas, are the most common class of models used to specify spatial random effects.
Under these Markov assumptions, observations made in areas close together are correlated, and more distant relationships are ignored.
Arguably the simplest GMRF model is that of @besag1991bayesian, where information is borrowed equally from each adjacent area.
This model is attractive as it requires minimal additional choices and is widely implemented.
As such, it has seen widespread use, including in agriculture [@oliver2015soil], ecology [@saracco2010modeling], epidemiology [@lawson2013statistical], image analysis [@schmid2006bayesian], neuroscience [@gossl2001bayesian] and public health [@dwyer2015drinking].
However, for irregular geometries like the administrative divisions of a country, the assumptions made by the Besag model about space are unrealistic.
As such, we set out to test the hypothesis that using more realistic assumptions about spatial structure would improve the performance of small-area estimation models.
In doing so, we provide practical recommendations and further understanding of models for spatial structure.
<!-- @gutreuter2019improving use the SAR model, which is still adjacency based but is not exactly the Besag model--restructure introduction to be more general and include introduction of other adjacency based models (which all have the same criticisms)? -->

@@

## Areal and point spatial data

Let $\Sc \subset \mathbb{R}^2$ be the study region, and the disjoint areas $\{A_i\}_{i = 1}^n$ be a spatial partition of $\Sc$.
Areal data are a type of spatial data where observations $y_i$ are associated to areas $A_i$.
Examples include the colour of a pixel, the minimum wage in a state, and the number of disease cases in a region.
Point data are another type of spatial data where instead observations $y(s)$ can be made at any location $s \in \Sc$ of a spatially-continuous stochastic process.
Areal data is often aptly conceptualised as arising from aggregation of point data, such that $y_i = \int_{A_i} y(s) \text{d}s$.
Exceptions include policies determined at an administrative level, such as the minimum wage or disease control measures, which have a genuinely discrete spatial structure.
<!-- Note that, for data of this sort, there is less clear justification for using spatial smoothing. -->
<!-- Realistically, data relate to many covariates, some number of which are likely have discrete spatial structure and others of which continuous spatial structure, such that the result is often an intermediate. -->
<!-- It may be possible to model this explicitly, though the result might look similar to what is achieved using e.g. BYM2. -->

If the areal data we observe indeed arise from aggregation, we may not use the underlying point data because (a) it is unobserved, (b) it may be inaccessible due to privacy constraints, administrative practicality or storage capacity, and (c) the point data are available, but we decide not to use them in the model.
In the final case (c), researchers may prefer to use area-level models as models for point data are typically more complex, require more data, are less immediately applicable to area-level policy making, and have higher computational burden.
Furthermore, seemly point data may actual be areal, such as data observed at polling stations or health facilities which individuals from the surrounding area travel to.
As such, choosing to model the underlying data generating mechanism as either areal or point is often a matter of pragmatism.

@@

Auto-correlation is an important property of most spatial processes.
Usually, outcomes for locations close together in space tend to be more similar than those far apart.
This fact is known as "Tobler's first law of geography" [@tobler1970computer] and, from a statistical point of view, is both a challenge and a benefit of working with spatial data.
Each observation provides less information than it would have had the samples been independent, making it more difficult to estimate global parameters.
However, spatial correlation can be used to improve local (indexed by a particular spatial location) parameter estimates particularly in parts of the study region where little to no information is available.

The latter benefit is basis for the statistical task of small-area estimation, which aims to produce reliable local estimates where small sample sizes lead to noisy data [@pfeffermann2013new].
In a spatial setting, this is often in small geographic areas, though the phrase "small-area" is not restricted to geographic areas and may be interpreted more broadly to mean any area where data are insufficient to make accurate local parameter inferences.
In the context of multilevel regression and post-stratification [@gelman1997poststratification], small-areas are generated by the intersection of demographic variables like age, gender and race, alongside geographic variables.
Due to the cost of gathering samples, a survey may be designed to give reliable estimates at at an aggregated level but not at a small-area level.
Although direct estimators of local parameters are unbiased, when data are sparse the total error may be reduced by accepting some bias in exchange for reduced variance using so-called indirect estimators.
Smoothing approaches use information from similar units to "borrow strength" from one parameter to another, with determining precisely what is meant by "similar" a central challenge.
For a recent review of both design-based and model-based approaches to spatial analysis of health-indicators, including both area and unit-level analysis, see @wakefield2020small.

@@

Bayesian hierarchical models are an attractive framework for small-area estimation, whereby areal data $\y = (y_1, \ldots, y_n)$ corresponding to the areas $\{A_i\}_{i = 1}^n$, are modelled using a three-layer structure [@berliner1996hierarchical; @cressie2015statistics; @rao2015small] given by
\begin{alignat}{2}
&\text{(Observations)}  &        y_i &\sim p(y_i \, | \, \x, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\
&\text{(Latent field)}  &        \x &\sim p(\x \, | \, \btheta), \label{eq:process} \\
&\text{(Parameters)}    & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
\end{alignat}
where $\x$ is a $N$-dimensional latent field, $\btheta$ are the $m$-dimensional parameters, $m < N$.
Spatial structure is modelled at the level of the latent field, and the observation model, typically either binomial or Poisson, is conditionally independent and identically distributed $p(y \, | \, x, \theta) = \prod_{i = 1}^n p(y_i \, | \, x_i, \theta)$.
As small-area estimation models are descriptive rather than mechanistic, it is natural [@best2005comparison] to use a flexible and computationally efficient Gaussian distribution $\x \sim \mathcal{N}(\bmu, \bSigma)$ for the latent field.
Three-layer models with Gaussian latent fields comprise a wide array of popular models, including generalised linear mixed models (GLMMs), and have been collectively studied under the title latent Gaussian models (LGMs) [@rue2009approximate].

@@

The latent field in a LGM is comprised a constant mean $\mu$, known also as the fixed effects, and spatial random effects $\phi \sim \mathcal{N}(0, \Sigma)$.
The mean may be modelled using a linear predictor $\mu_i = \beta_0 + z_i^\top \beta$, where $(\beta_0, \beta)$ are regression parameters corresponding to areal covariates $z_i = (z_{i,1}, \ldots, z_{i,p})^\top$.
We focus on the spatial random effects $\phi$, whose role is to capture the spatial structure between areas not already accounted for in the mean.
If no spatial structure remains after inclusion of covariates in the mean, independent and identically distributed (IID) random effects $\phi_i \sim \N (0, \tau_\phi^{-1})$ should be used, where $\tau_\phi$ is a shared precision.
Exploratory data analysis techniques such as visual inspection or Moran's $I$ coefficient [@cliff1981spatial] may be used to determine if there remains any spatial auto-correlation in the data.
<!-- TODO: Add calculation of Moran's $I$ to my HIV survey data-sets. -->
Specifying $\phi$ amounts to specifying the entries of a precision or covariance matrix, as we discuss in the following sections.

@@

Spatial structure can be encoded using a symmetric relation between areas, where $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring.
Adjacency is often defined by a shared border, though other choices, such as inclusion of second or higher-degree neighbours [@paciorek2013spatial], are also possible.
The Besag model [@besag1991bayesian] is a type of improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \phi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\phi_{-i} = (\phi_1, \ldots, \phi_{i - 1}, \phi_{i + 1}, \ldots, \phi_n)^\top$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \phi \sim \mathcal{N}(0, \tau_\phi^{-1} R^{-}), \label{eq:gmrf}
\end{equation}
where $R^{-}$ is the generalised inverse of the rank-deficient structure matrix $R$, so-called because it defines the \textit{structure} of the precision matrix, with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \phi_{-i}) = p(\phi_i \, | \, \phi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $R$ whereby $\phi_i \perp \phi_j \, | \, \phi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $R$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:geometry-graph-zwe} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
Shortly, in Section \ref{sec:concerns} we discuss the appropriateness of using the graph rather than the complete geometry.

Rewriting Equation \ref{eq:gmrf}, the probability density function of $\phi$ is
\begin{equation}
    p(\phi)
    \propto \exp \left( -\frac{\tau_\phi}{2} \phi^\top R \phi \right)
    \propto \exp \left( -\frac{\tau_\phi}{2} \sum_{i \sim j} (\phi_i - \phi_j)^2 \right). \label{eq:pdfphi}
\end{equation}
This density is a function of the pairwise differences $\phi_i - \phi_j$ and so is invariant to the addition of a constant $c$ to each entry $p(\phi) = p(\phi + c1)$, leading to an improper uniform distribution on the average of the $\phi_i$.
If $\mathcal{G}$ is connected, in that by traversing the edges, any vertex can be reached from any other vertex, then there is only one impropriety in the model and $\text{rank}(R) = n - 1$, while if $\mathcal{G}$ is disconnected, and composed of $n_c \geq 2$ connected components with index sets $I_1, \ldots, I_{n_c}$, then the corresponding structure matrix $R$ has rank $n - n_c$ and the density is invariant to the addition of a constant to each of the connected components $p(\phi_{I}) = p(\phi_{I} + c1)$ where $I = I_1, \ldots, I_{n_c}$.

@@

\begin{figure}
\centering
\includegraphics{depends/maup1.pdf}
\caption{Each of these four geometries corresponds to the same adjacency graph.}
\label{fig:tikz1}
\end{figure}

@@

The Besag model was originally proposed for use in image analysis, where areas correspond to pixels arranged in a regular lattice structure.
Since then, it has seen wider use, including in situations where the spatial structure is less regular.
There are a number of concerns about the model's applicability to this broader setting, which we highlight below.
Closely linked are the modifiable areal unit problem [@openshow1979million] whereby statistical conclusions change as a result of seemingly arbitrary changes in data aggregation, and the challenge of ecological inference and the ecological fallacy [@wakefield2010aggregation], which applies particularly to inference about individuals and groups.
<!-- Missing an item about potential loss of information by using a sparse matrix, though I don't actually think this is very important. -->

@@

Summarising all spatial information with an adjacency graph represents a loss of information.
Many geometries share the same adjacency graph, and therefore the same probability model, as illustrated by Figure \ref{fig:tikz1}.
This fact is not concerning in itself, though it prompts consideration as to whether the class of geometries with the same adjacency graph is sufficiently similar to merit identical models.\footnote{The regularity of realistic geometries may help to constrain each class to be more similar than it strictly has to be. In other words, although pathological geometries can be constructed, they are implausible in statistical practise and so not of great concern to us here.}
It is intuitive that the more regular the spatial structure the less information is lost in compression to an adjacency graph.
In image analysis, very little spatial information is lost in compression a lattice structure to an adjacency graph.
On the other hand, the regions of a country, determined by political and geographic forces, tend to display greater irregularity.
As such, the appropriateness of adjacency compression varies by the type of geometry common to the application setting.

@@

A feature specific to the Besag model is that all adjacent areas count equally.
This assumption is unsatisfying: for most geometries, we expect some areas to be more highly correlated to some neighbours than to others.
Figure \ref{fig:tikz1} illustrates a number of heuristic features for neighbour importance, including length of shared border in geometry (iii), and the proximity of centers of mass in geometry (iv).
This may be remedied using more general weighted ICAR models, as we outline in Section \ref{sec:icar}.

@@

A striking feature of Equation \ref{eq:besag} is that the precision of $\phi_i$ is proportional to its number of neighbours $n_{\delta i}$.
It follow that as $n_{\delta i} \to \infty$ then $\text{Var}(\phi_i) \to 0$.
This is illustrated by Figure \ref{fig:tikz2} where the area on the right is repeatedly divided such that its number of neighbours increases by one at each step.
This property is a consequence of averaging the conditional mean over a greater number of areas, which, in certain situations, can correspond to a greater amount of information.
However, if the amount of information in the shaded area remains fixed, it is inappropriate that $\text{Var}(\phi_1)$ should tend to zero just as a result of drawing additional, arbitrary, boundaries.
In the image analysis setting this modelling assumption is reasonable: each pixel represents a fixed amount of information and a higher pixel density represents a greater amount of information.
On the other hand, in public health and epidemiology, drawing boundaries to create additional regional areas cannot be expected to correspond to a greater amount of information and this assumption is not appropriate.

@@

The Besag model is a special case of a more general class of (zero-mean) ICAR models in which each neighbour may not be weighted equally.
These models can be specified in terms of scaled weights $\{b_{ij}\}_{j \sim i}$ and a precision vector $\kappa = (\kappa_1, \ldots, \kappa_n)^\top$ such that $\phi_i \, | \, \phi_{-i} \sim \N ( \sum_{j: j \sim i} b_{ij} \phi_j, (\kappa_i \tau_\phi)^{-1} )$.
The structure matrix $R$ corresponding to the above full conditionals is given by $R = D_\kappa(I - B)$, where $B$ has the entries $b_{ij}$ for $i \sim j$, diagonal entries $b_{ii} = 0$ and $b_{ij} = 0$ for $i \nsim j$ and the matrix $D_\kappa$ is given by $\text{diag}(\kappa_1, \ldots, \kappa_n)$.
As the structure matrix is symmetric we must have the condition $-b_{ij}\kappa_i = -b_{ji}\kappa_j$.
To meet this condition, it is often simpler to use an unscaled weights matrix $W = D_\kappa B$ such that $R = D_\kappa - W$.
For example, in the Besag model $W$ corresponds to the adjacency matrix.
The scaled weights can then be recovered by $b_{ij} = w_{ij} / \kappa_i$ where $\kappa_i = \sum_{k: k \sim i} w_{ik}$.
A thorough discussion of methods for specifying $W$ is provided by @duncan2017spatial.
Much of the work in this area focuses on the case where the geometry is a lattice.

@@

# Acknowledgements {-}

>>>>

---
title: Understanding models for spatial structure in small-area estimation
classoption: ba
author:
  - firstname: Adam 
    lastname: Howes
    email: ath19@ic.ac.uk
    affiliationref: A,B
    footnoterefs: 
      - t1
  - firstname: Jeffrey W.
    lastname: Eaton
    affiliationref: C,B
  - firstname: Seth R.
    lastname: Flaxman
    affiliationref: D
affiliations:
  - ref: A
    name: Imperial College London 
    address: Department of Mathematics
  - ref: B
    name: Imperial College London
    address: MRC Centre for Global Infectious Disease Analysis
  - ref: C
    name: Harvard University
    address: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health
  - ref: D
    name: University of Oxford
    address: Department of Computer Science 
footnotes:
  - ref: t1
    text: "Corresponding Author"
supplements:
  - title: Title of Supplement A
    description: Short description of Supplement A.
  - title: Title of Supplement B
    description: Short description of Supplement B.
acknowledgements: |
  | Yidan Xu undertook the Mary Lister McCammon Summer Research Fellowship 2019 on related work. AH thanks Robert Ashton for help with \textsf{R} package development, and Harrison Zhu, Theo Rashid, Samir Bhatt, Tim Lucas, Elizaveta Semenova, Marta Blangiardo and Oliver Ratmann for helpful conversations.
abstract: |
  |  Spatial correlation between small-areas is typically accounted for using spatial random effects. However, commonly used models for spatial random effects based upon adjacency relations, like the Besag model, make unrealistic and unsatisfying spatial assumptions for non-grid geometries. Furthermore, they do not take into account that across many settings areal data is generated by aggregating continuous geostatistical data. Defining spatial random effects which instead correspond to aggregated Gaussian processes gives a more intuitively convincing correlation structure, and allows tools from kernel methods to be used in models for areal data.
  |
  |  In a simulation study, we used a proper scoring rule to evaluate the performance of different spatial random effect specifications. The simulations were performed under model misspecification and across a range of vignette and realistic geometries with different regularities. We also compared model performance in estimating district-level HIV prevalence from PHIA household survey data across four countries in sub-Saharan Africa, using cross-validation and spatial cross-validation.
date: "`r Sys.Date()`"
bibliography: citations.bib
#linenumbers: true
#numbersections: true
output: 
  bookdown::pdf_book: 
    base_format: rticles::isba_article
    md_extensions: -autolink_bare_uris #keep these off or \printead{} will fail on email addresses
    includes:
      in_header: preamble.tex
---

@@

Spatial random effects are frequently used to model spatial variation in areal data [@haining2003spatial; @cramb2018investigation].
The most common class of models used to specify spatial random effects are Gaussian Markov random fields (GMRFs), which combine a Gaussian distribution with Markov conditional independence assumptions between areas [@rue2005gaussian].
Under the Markov assumptions, observations made in areas close together are correlated, and more distant relationships are ignored.
Perhaps the simplest GMRF model is that of @besag1991bayesian, where information is borrowed equally from each adjacent area.
This model is attractive as it requires minimal additional choices and is widely implemented.
As such, it has seen widespread use, including in agriculture [@oliver2015soil], ecology [@saracco2010modeling], epidemiology [@lawson2013statistical], image analysis [@schmid2006bayesian], neuroscience [@gossl2001bayesian] and public health [@dwyer2015drinking].
However, for irregular geometries like the administrative divisions of a country, the assumptions made by the Besag model about space are unrealistic.
As such, in this work we test the hypothesis that using more realistic assumptions about spatial structure improves the performance of small-area estimation models.
In doing so, we provide practical recommendations and further understanding of models for spatial structure.
<!-- @gutreuter2019improving use the SAR model, which is still adjacency based but is not exactly the Besag model--restructure introduction to be more general and include introduction of other adjacency based models (which all have the same criticisms)? -->

@@

## Areal and geostatistical spatial data

Let $\Sc \subset \mathbb{R}^2$ be the study region, and the disjoint areas $\{A_i\}_{i = 1}^n$ be a spatial partition of $\Sc$ such that $A_i \cap A_j = \emptyset$ and $\cup_{i = 1}^n A_i = \Sc$.
Areal data are a type of spatial data where observations $y_i \in \mathcal{Y}$ are associated to areas $A_i$.
Examples of areal data include the colour of a pixel, the minimum wage in a state, and the number of disease cases in a region.
Geostatistical data are another type of spatial data where instead observations $y(s) \in \mathcal{Y}$ can be made at any location $s \in \Sc$ of a spatially-continuous stochastic process.
Examples of geostatistical spatial data include the temperature at a monitoring station, the response to a household survey, and the quality of a soil sample.

Areal data may often be conceptualised as arising from aggregation of geostatistical data, such that $y_i = \int_{A_i} y(s) \text{d}s$.
Notable exceptions include policies determined at an administrative level, such as the minimum wage or disease control measures, which have a truly discrete spatial structure.
<!-- Note that, for data of this sort, there is less clear justification for using spatial smoothing. -->
<!-- Realistically, data relate to many covariates, some number of which are likely have discrete spatial structure and others of which continuous spatial structure, such that the result is often an intermediate. -->

In some situations it may be preferable to work using aggregated areal data rather than the original geostatistical data.
For one, the geostatistical data may be unavailabile or inaccessible due to privacy constraints, administrative practicality or limitations in storage capacity.
Alternatively, opting to model at the area-level can be advantageous as models for geostatistical data tend to be more complex, require more data, have higher computational burden, and offer less immediate applicablility to area-level policy-making.
In addition, in some circumstances, seemingly geostatistical data may truly be areal, such as data observed at polling stations or health facilities which individuals from the surrounding area travel to.
Ultimately, the decision to model the underlying data generating mechanism as either areal or geostatistical is often a matter of practical considerations and pragmatism.

@@

Auto-correlation is an important property of most spatial processes.
Usually, outcomes for locations close together in space tend to be more similar than those far apart, known as "Tobler's first law of geography" [@tobler1970computer].
This property both presents a challenge and an opportunity.
Each observation provides less information than it would have had the samples been independent, making it more difficult to estimate global parameters.
On the other hand, spatial correlation can be used to improve local parameter estimates, indexed by a particular spatial location, especially in parts of the study region where little to no information is available.

The latter benefit is basis for the statistical task of small-area estimation, which aims to produce reliable local estimates where small sample sizes lead to noisy data [@pfeffermann2013new].
In a spatial setting, this is often in small geographic areas. 
However, the phrase "small-area" is not restricted to geographic areas and may be interpreted more broadly to mean any area where data are insufficient to make accurate local parameter inferences.
For example, in the context of multilevel regression and post-stratification [@gelman1997poststratification], small-areas are generated by the intersection of demographic variables like age, gender and race, alongside geographic variables.
Because of the cost of gathering samples, a survey may be designed to give reliable estimates at at an aggregate level but not at a small-area level.
Although direct estimators of local parameters are unbiased, when data are sparse the total error may be reduced by accepting some bias in exchange for reduced variance using so-called indirect estimators.
Smoothing approaches use information from similar units to "borrow strength" from one parameter to another, with determining precisely what is meant by "similar" a central challenge.
See @wakefield2020small for a recent review of both design-based and model-based approaches to small-area estimation of health-indicators, including both area and unit-level analysis.

@@

Bayesian hierarchical models are an attractive framework for small-area estimation.
Let $[n]$ denote $1, \ldots, n$.
Areal data $\y = (y_i)_{i \in [n]}$ may be modelled using a three-layer structure [@berliner1996hierarchical; @cressie2015statistics; @rao2015small] given by
\begin{alignat}{2}
&\text{(Observations)}  &        \y &\sim p(\y \, | \, \x, \btheta_1), \label{eq:data} \\
&\text{(Latent field)}  &        \x &\sim p(\x \, | \, \btheta_2), \label{eq:process} \\
&\text{(Parameters)}    & \qquad (\btheta_1, \btheta_2) &= \btheta \sim p(\btheta), \label{eq:parameters}
\end{alignat}
where $\x = (x_i)_{i \in [N]}$ is a $N$-dimensional latent field, and $\btheta = (\theta_i)_{i \in [m]}$ are the $m$-dimensional hyperparameters, with $m < N$.
Small-area estimation models are usually descriptive rather than mechanistic, so a natural choice is to model the latent field with a flexible and computationally efficient Gaussian distribution $\x \sim \mathcal{N}(\bmu, \mathbf{Q})$ with mean vector $\bmu$ and precision matrix $\mathbf{Q}$.
Three-layer models with Gaussian latent fields have been collectively studied as latent Gaussian models (LGMs) [@rue2009approximate] and comprise a wide array of popular models, including generalised linear mixed models.

@@

The observations in an LGM are conditionally independent and identically distributed given linear predictors $\bmeta = (\eta)_{i \in [n]}$ such that
\begin{equation*}
p(\y \, | \, \, \x, \btheta_1) = p(\y \, | \, \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1).
\end{equation*}
Each linear predictor $\eta_i$ is constructed from the Gaussian latent field such that
\begin{equation*}
\eta_i = \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{equation*}
where the intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and functions $f_k(\cdot)$ of the covariates $u_{ki}$ are each Gaussian and may be collected into $\x$.
The conditional mean $\mathbb{E}(y_i \, | \, \eta_i)$ is given by $g(\eta_i)$ where $g: \mathbb{R} \to \mathcal{Y}$ is an inverse link function

Spatial random effects $\phi(\cdot)$ are one Gaussian function which may be included in the model, whose role is to capture the spatial structure between areas not already accounted for by $\beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji}$.
Should no spatial structure remain, then independent and identically distributed (IID) random effects $\phi_i \sim \N (0, \tau_\phi^{-1})$ may be appropriate, where $\tau_\phi$ is a shared precision.
Exploratory data analysis techniques such as visual inspection or Moran's $I$ coefficient [@cliff1981spatial] may be used to determine if there remains any spatial auto-correlation in the data.
<!-- TODO: Add calculation of Moran's $I$ to my HIV survey data-sets. -->
Specifying $\bphi = (\phi_i)_{i \in [n]}$ amounts to specifying the entries of a precision or covariance matrix, as we discuss in the following sections.

@@

Spatial structure can be encoded using a symmetric relation between areas.
Let $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring.
Adjacency is often defined by a shared border, though other choices are also possible [@paciorek2013spatial].
The Besag model [@besag1991bayesian] is an improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \bphi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\bphi_{-i}$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \bphi \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} \bR^{-}), \label{eq:gmrf}
\end{equation}
where $\bR^{-}$ is the generalised inverse of the rank-deficient structure matrix $\bR$, so-called because it defines the structure of the precision matrix, with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \bphi_{-i}) = p(\phi_i \, | \, \bphi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $\bR$ whereby $\phi_i \perp \phi_j \, | \, \bphi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $\bR$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:geometry-graph-zwe} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
In Section \ref{sec:concerns} we will discuss the appropriateness of using the graph rather than the complete geometry.

Rewriting Equation \ref{eq:gmrf}, the probability density function of $\bphi$ is
\begin{equation}
    p(\bphi)
    \propto \exp \left( -\frac{\tau_\phi}{2} \bphi^\top \bR \bphi \right)
    \propto \exp \left( -\frac{\tau_\phi}{2} \sum_{i \sim j} (\phi_i - \phi_j)^2 \right). \label{eq:pdfphi}
\end{equation}
This density is a function of the pairwise differences $\phi_i - \phi_j$ and so is invariant to the addition of a constant $c$ to each entry $p(\bphi) = p(\bphi + c\mathbf{1})$, leading to an improper uniform distribution on the average of the $\phi_i$.
If $\mathcal{G}$ is connected, in that by traversing the edges, any vertex can be reached from any other vertex, then there is only one impropriety in the model and $\text{rank}(\bR) = n - 1$, while if $\mathcal{G}$ is disconnected, and composed of $n_c \geq 2$ connected components with index sets $I_1, \ldots, I_{n_c}$, then the corresponding structure matrix $\bR$ has rank $n - n_c$ and the density is invariant to the addition of a constant to each of the connected components $p(\bphi_{I}) = p(\bphi_{I} + c\mathbf{1})$ where $I = I_1, \ldots, I_{n_c}$.

@@

\begin{figure}
\centering
\includegraphics{depends/maup1.pdf}
\caption{Each of these four geometries have the same adjacency graph.}
\label{fig:tikz1}
\end{figure}

@@

The Besag model was originally proposed for use in image analysis, where areas correspond to pixels arranged in a regular lattice structure.
Since then, it has seen wider use, including in situations where the spatial structure is less regular.
We highlight a number of concerns about the model's applicability to this broader setting.
This discussion is closely linked to the modifiable areal unit problem [@openshow1979million] whereby statistical conclusions change as a result of seemingly arbitrary changes in data aggregation, and the challenge of ecological inference and the ecological fallacy [@wakefield2010aggregation], which applies particularly to inference about individuals and groups.
<!-- Missing an item about potential loss of information by using a sparse matrix, though I don't actually think this is very important. -->

@@

Summarising all spatial information with an adjacency graph represents a loss of information.
Many geometries share the same adjacency graph, and therefore the same probability model, as illustrated by Figure \ref{fig:tikz1}.
This fact is not concerning in itself, but it prompts consideration as to whether the class of geometries with the same adjacency graph is sufficiently similar to merit identical models.^[The regularity of realistic geometries may help to constrain each class to be more similar than it strictly has to be. In other words, although pathological geometries can be constructed, they are implausible in statistical practise and so not of great concern to us here.]
Intuitively the more regular the spatial structure, the less information is lost in compression to an adjacency graph.
In image analysis, very little spatial information is lost in compression a lattice structure to an adjacency graph.
On the other hand, the regions of a country, determined by political and geographic forces, tend to display greater irregularity.
As such, the appropriateness of adjacency compression varies by the type of geometry common to the application setting.

@@

In the Besag model is all adjacent areas count equally.
This assumption is unsatisfying: for most geometries, we expect some areas to be more highly correlated to some neighbours than to others.
Figure \ref{fig:tikz1} illustrates a number of heuristic features for neighbour importance, including length of shared border, and the proximity of centers of mass.
This may be remedied using more general weighted ICAR models, as we outline in Section \ref{sec:icar}.

@@

In Equation \ref{eq:besag} the precision of $\phi_i$ is proportional to its number of neighbours $n_{\delta i}$.
It follow that as $n_{\delta i} \to \infty$ then $\text{Var}(\phi_i) \to 0$.
This is illustrated by Figure \ref{fig:tikz2} where the area on the right is repeatedly divided such that its number of neighbours increases.
This property is a consequence of averaging the conditional mean over a greater number of areas, which, in certain situations, can correspond to a greater amount of information.
However, if the amount of information in the shaded area remains fixed, it seems inappropriate that $\text{Var}(\phi_1)$ should tend to zero simply as a result of drawing additional, arbitrary, boundaries.
In the image analysis setting this modelling assumption is reasonable: each pixel represents a fixed amount of information and a higher pixel density represents a greater amount of information.
On the other hand, in public health and epidemiology, drawing boundaries to create additional areas is not expected to correspond to a greater amount of information.

@@

The Besag model is a special case of a more general class of (zero-mean) ICAR models in which each neighbour may not be weighted equally.
These models can be specified in terms of scaled weights $\{b_{ij}\}_{j \sim i}$ and a precision vector $\bkappa = (\kappa_i)_{i \in [n]}$ such that $\phi_i \, | \, \bphi_{-i} \sim \N ( \sum_{j: j \sim i} b_{ij} \phi_j, (\kappa_i \tau_\phi)^{-1} )$.
The structure matrix $\bR$ corresponding to the above full conditionals is given by $\bR = \mathbf{D}_\kappa(\mathbf{I} - \mathbf{B})$, where $\mathbf{B}$ has the entries $b_{ij}$ for $i \sim j$, diagonal entries $b_{ii} = 0$ and $b_{ij} = 0$ for $i \nsim j$ and the matrix $\mathbf{D}_\kappa$ is given by $\text{diag}(\kappa_1, \ldots, \kappa_n)$.
As the structure matrix is symmetric we must have the condition $-b_{ij}\kappa_i = -b_{ji}\kappa_j$.
To meet this condition, it is often simpler to use an unscaled weights matrix $\mathbf{W} = \mathbf{D}_\kappa \mathbf{B}$ such that $\mathbf{R} = \mathbf{D}_\kappa - \mathbf{W}$.
For example, in the Besag model $W$ corresponds to the adjacency matrix.
The scaled weights can then be recovered by $b_{ij} = w_{ij} / \kappa_i$ where $\kappa_i = \sum_{k: k \sim i} w_{ik}$.
A thorough discussion of methods for specifying $\mathbf{W}$ is provided by @duncan2017spatial.
Much of the work in this area focuses on the case where the geometry is a lattice.

@@
